---
title: "World Happiness Index"
author: "Nguyen Ngo, Jenny Tong, Ava Walters"
date: "12/02/2025"
format: 
  pdf:
    documentclass: article
    # margin: 1 in all sides
    geometry:
      - margin=1in
    
    # no table of content
    toc: false
    
    # figures with captions on top
    fig-pos: "H"
    tbl-cap-location: top
    fig-cap-location: top
 
# prevent warnings and messages
execute:
  echo: false
  warning: false
  message: false
  error: false
---

# Introduction

This report contains statistical analysis on ...

1.  **Life Evaluation vs Rank:** Citizens' life satisfaction correlation with their country's happiness ranking.
2.  **Economic Influence:** The relationship between GDP per capita and happiness index.
3.  **Governance Impact:** Corruption perception effect happiness rankings.

This analysis employs Exploratory Data Analysis techniques to find different relationships in data and provide insights into what contribute to a nation's happiness index.

# Data Provenance and Ethical Considerations (??) {#sec-armed}

## FAIR (?)

## CARE (?)

## Data Preparation

\vspace{8mm}

```{r world-happiness-index-wrangling, include=FALSE}
 # Wrangling Code that will not be shown when rendered

# STEP 1: Load required pacakges
library(tidyverse)
library(readxl) # read excel file
library(janitor)

# STEP 2: Load and clean the Excel dataset
happiness_raw <- read_excel(
  "/Applications/PSU/FALL 2025/STAT 184/week 12/nguyen_repo_practice/Sec-3_FP_NguyenNgo_AvaWalters_JennyTong/World Happiness Report Data Figure 2.1.xlsx",
sheet = "Data for Figure 2.1 (2011–2024)"
) %>%
  # standardize column names
  clean_names()

# STEP 3: Filter for 2024 data
# select important columns
happiness_2024 <-  happiness_raw %>% 
  filter (year == 2024) %>% 
  # rename columns
  select(
    country = country_name,
    rank,
    life_evaluation = life_evaluation_3_year_average,
    gdp = explained_by_log_gdp_per_capita,
    social_support = explained_by_social_support,
    health_life_expectancy = explained_by_healthy_life_expectancy,
    freedom = explained_by_freedom_to_make_life_choices,
    generosity = explained_by_generosity,
    corruption = explained_by_perceptions_of_corruption,
    dystopia_residual
  ) %>% 
  # fix incomplete rows and convert ranks to integers
  mutate(
    rank = as.integer(rank),
    across(c(life_evaluation:corruption), as.numeric)  
  ) %>% 
  filter(!is.na(rank), !is.na(life_evaluation))

# STEP 4: Calculate the Sensitivity Metrics
happiness_sensitivity <- happiness_2024 %>% 
  # order by descending life evaluation
  arrange(desc(life_evaluation)) %>% 
  mutate(
    life_eval_diff = life_evaluation - lead(life_evaluation), 
    rank_diff = rank - lead(rank),
    sensitivity = abs(rank_diff / life_eval_diff), # measures sensitivity
    sensitivity_group = cut(sensitivity,
                            breaks = c(0, 5, 10, 20, Inf),
                            Labels = c("Low", "Medium", "High", "Very High")
                            )
  ) %>% 
  filter(!is.na(sensitivity))

# STEP 5: Compute Correlation of Factors w/ Rank
factor_correlations <- happiness_2024 %>%
select(rank, life_evaluation, gdp, corruption, social_support,
health_life_expectancy, freedom, generosity) %>% 
cor(use = "complete.obs") %>% # compute correlation matrix
as.data.frame() %>%
rownames_to_column("factor") %>% # convert rownames to a column
select(factor, correlation_with_rank = rank)

# STEP 6: Top 10 Happiest Countries Table
top_10 <- happiness_2024 %>% 
  # select 10 smallest ranks
  slice_min(rank, n = 10) %>% 
  select(country, rank, life_evaluation, gdp, corruption)

# STEP 7: Create Bottom 10 Least Happiest Countries Table
bottom_10 <- happiness_2024 %>% 
  # select 10 largest ranks
  slice_max(rank, n = 10) %>% 
  select(country, rank, life_evaluation, gdp, corruption)

# STEP 8: A Summary Statistics Table, if needed
summary_stats <- happiness_2024 %>% 
  summarize(
    # count countries
    n_countries = n(),
    avg_life_eval = mean(life_evaluation, na.rm = TRUE),
    avg_gdp = mean(gdp, na.rm = TRUE),
    avg_corruption = mean(corruption, na.rm = TRUE),
    life_eval_range = max(life_evaluation) - min(life_evaluation),
    rank_range = max(rank) - min(rank)
  )
```

Explanation

```         
```

\vspace{8mm}

## Summary Statistics (? maybe to display what we have first then go into tables and figures)

# Visualization

## 1. Life Evaluation vs Country Rank

\vspace{4mm}

```{r}

#| tbl-cap: "Life Evaluation Statistics by Rank Group"

# Create rank groups first
happiness_2024 <- happiness_2024 %>% 
  mutate(
    rank_group = cut(rank, 
                     breaks = quantile(rank, probs = seq(0, 1, 0.2)),
                     labels = c("Top 20%", "20-40%", "40-60%", "60-80%", "Bottom 20%"),
                     include.lowest = TRUE)
  )

# Create the summary table
life_eval_summary <- happiness_2024 %>% 
  group_by(`Rank Group` = rank_group) %>% 
  summarize(
    `Number of Countries` = n(),
    `Avg. Life Evaluation` = round(mean(life_evaluation, na.rm = TRUE), 2),
    `Min` = round(min(life_evaluation, na.rm = TRUE), 2),
    `Max` = round(max(life_evaluation, na.rm = TRUE), 2),
    `STDEV` = round(sd(life_evaluation, na.rm = TRUE), 2)
  ) %>% 
  arrange(`Rank Group`)

knitr::kable(life_eval_summary)
```

\vspace{4mm}

```{r}
#| label: fig-life-eval-vs-rank
#| fig-cap: "Life Evaluation vs Country Rank"
#| fig-alt: "Scatter plot showing life evaluation scores on x-axis and country ranks on y-axis. Points are color-coded by rank groups."

ggplot(happiness_2024, aes(x = life_evaluation, y = rank)) +
  geom_point(aes(color = rank_group), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue", linewidth = 1) +
  scale_y_reverse() +  # Reverse so rank 1 is at top
  scale_color_brewer(palette = "Set1", name = "Rank Group") +
  labs(
    title = "Life Evaluation vs Country Rank (2024)",
    x = "Life Evaluation Score (0-10 scale)",
    y = "Rank (1 = Happiest)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),  # Move x-axis label down
    axis.title.y = element_text(margin = margin(r = 10)),  # Move y-axis label right
    legend.position = "right",
    plot.margin = margin(20, 20, 20, 20)  # Add margin around entire plot
  )
```

**Interpretation**:

## 2. GDP Contribution vs Country Rank

\vspace{4mm}

```{r}
#| tbl-cap: "GDP Contribution Statistics by Rank Group"

# Create GDP summary table
gdp_summary <- happiness_2024 %>% 
  group_by(`Rank Group` = rank_group) %>% 
  summarize(
    `Number of Countries` = n(),
    `Avg. GDP Contribution` = round(mean(gdp, na.rm = TRUE), 2),
    `Min` = round(min(gdp, na.rm = TRUE), 2),
    `Max` = round(max(gdp, na.rm = TRUE), 2),
    `STDEV` = round(sd(gdp, na.rm = TRUE), 2)
  ) %>% 
  arrange(`Rank Group`)

knitr::kable(gdp_summary)
```

\vspace{4mm}

```{r}
#| label: fig-gdp-vs-rank
#| fig-cap: "GDP Contribution vs Country Rank"
#| fig-alt: "Scatter plot showing GDP contribution scores on x axis and country ranks on y axis. Points are color-coded by rank groups."

ggplot(happiness_2024, aes(x = gdp, y = rank)) +
  geom_point(aes(color = rank_group), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "darkblue", linewidth = 1) +
  scale_y_reverse() +
  scale_color_brewer(palette = "Set1", name = "Rank Group") +
  labs(
    title = "GDP Contribution vs Country Rank (2024)",
    x = "GDP Contribution Score (Log GDP per capita)",
    y = "Rank (1 = Happiest)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14, margin = margin(b = 15)),
    axis.title.x = element_text(margin = margin(t = 10)),  # move x-axis label down
    axis.title.y = element_text(margin = margin(r = 10)),  # move y-axis label right
    legend.position = "right",
    plot.margin = margin(20, 20, 20, 20)  # add margin around entire plot
  )
```

**Interpretation**: The table and graph show that as GDP and rank have a relatively strong, positive, linear relationship. This means that as GDP increases then so will rank. This means that countries that have higher GDP are more likely to have a higher rank and be more well off. If a country is struggling to create world wide markets, new goods, and build capital, than their government and people are going to struggle. This means the output of the country will be lower and thus rank will reflect this. You want your country to have a high GDP!

## 3. Corruption Perception vs Country Rank

\vspace{4mm}

```{r}

```

\vspace{4mm}

```{r}

```

**Interpretation**:

# Discussion and Conclusions

## Key Findings

1.  **Life Evaluation**:

2.  **Economic Foundation**

3.  **Governance Importance**

## Limitations and Future Research

# Author Contributions

# References

# Code Appendix
